---
title: "Log link regression priors"
author: "Adam Smith"
date: "2022-12-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r packages, message=FALSE, warning=FALSE, out.width=8}
library(tidyverse)
library(rethinking)
library(kableExtra)
library(patchwork)
library(scales)

birds <- read.csv("bird_data_bayes_stats.csv")
```

## Priors in log-link regression models

When modeling counts, presence-absence or other data that are constrained to be non-negative, models often use a log-link. It is important to remember the link function when thinking about appropriate priors for parameters in the model. 


```{r data_setup}


sp_indicators <- birds %>% #creates a data frame that list each species indicator number
  filter(bird_guild == "Forest") %>% 
  select(english) %>% 
  distinct() %>% # selects just the unique rows
  mutate(species_ind = as.integer(factor(english))) %>% 
  arrange(species_ind)


# forest cover compare partial pool with no pooling ----------------------------------------------


birds_sub <- birds %>% 
  mutate(route_ind = as.integer(factor(RouteName)),
         #species_ind = as.integer(factor(english)),
         guild_ind = as.integer(factor(bird_guild)),
         foot_scale = scale(human_footprint),
         change_foot_scale = scale(change_human_footprint),
         pforest_scale = (proportion_forest-0.5)) %>% 
  inner_join(sp_indicators,
            by = c("english")) %>% 
  select(count,species_ind,pforest_scale) # this removes all unnecessary columns to keep ulam() happy

```



## Exploring priors in a simplified way

If the model is relatively simple, you can explore the priors distributions by transforming the priors (which are coded on the log-scale) into the original scale to see how they relate to the data.

For example, here are 4 alternate priors for the intercept terms in a regression model estimating the effect of the proportion of forest in the surrounding landscape on the abundance of forest birds. I've generated random draws from these priors, re-transformed them using the `exp()` function (`exp(log(a)) = a`), then added a rug-plot (dark lines along the x-axis) to indicate the observed means for each species.

```{r priors_1, message=FALSE, warning=FALSE}

obs_means <- birds_sub %>% 
  group_by(species_ind) %>% 
  summarise(mean = mean(count))

nn <- 10000#number of samples to draw from the prior
mean_priors = data.frame(prior = exp(c(rnorm(nn,0,1),
                                   rnorm(nn,3,1),
                                   rnorm(nn,0,0.5),
                                   rnorm(nn,0,5))),
                         intercept_prior = c(rep("Normal(0,1)",nn),
                                             rep("Normal(3,1)",nn),
                                             rep("Normal(0,0.5)",nn),
                                             rep("Normal(0,5)",nn)))
bks = seq(0,100,0.5)
prior_obs <- ggplot(data = mean_priors)+
  geom_freqpoly(aes(x = prior,colour = intercept_prior),breaks = bks,center = 0)+
  geom_rug(data = obs_means, aes(x = mean),inherit.aes = FALSE,
           size = 1,alpha = 0.5)+
  coord_cartesian(xlim = c(0,20))+
  xlab("Mean Counts")

prior_obs_zoom <- ggplot(data = mean_priors)+
  geom_freqpoly(aes(x = prior,colour = intercept_prior),breaks = bks,center = 0)+
  geom_rug(data = obs_means, aes(x = mean),inherit.aes = FALSE,
           size = 1,alpha = 0.5)+
  coord_cartesian(ylim = c(0,200),
                  xlim = c(0,20))+
  xlab("Mean Counts")

print(prior_obs + prior_obs_zoom + plot_layout(guides = "collect") 
      + plot_annotation(title = "Alternate intercept priors compared to the observed mean counts by species"))


```


So the priors with mean = 0 seem to capture the observed means relatively well, most mean counts are pretty small (< 3-5 birds). The prior with a positive mean value (`normal(3,1)`) doesn't seem to fit the data at all. This positive prior could seem perfectly intuitive if you had forgotten about the log-link in the model (counts are always positive), but it actually has relatively little prior probability mass at the most commonly observed mean counts (< 1) and has most of its prior probability mass at values greater than the highest observed mean count. If you studied birds a lot, you might also expect to see mean counts for some very common species that were on the order of 10 or more. In that case the `normal(0,0.5)` prior seems like it might be a little too narrow.



You can also combine multiple priors to explore the combined effects on the expected values.

```{r priors_2, message=FALSE, warning=FALSE}

#setting up a dataframe of forest cover values that span the available range 
# and have a few different scalings
forest_range <- data.frame(proportion_forest = seq(0,0.8,length.out = 100)) %>% 
  mutate(pforest_centered = proportion_forest-0.5,
         pforest_scaled = scale(proportion_forest))

nn <- 100#number of samples to draw from the prior
priors_changes = data.frame(intercept = c(rnorm(nn*3,0,1)),
                         slope = c(rnorm(nn,0,1),
                                   rnorm(nn,0,3),
                                   rnorm(nn,0,5)),
                         slope_prior = c(rep("Normal(0,1)",nn),
                                             rep("Normal(0,3)",nn),
                                             rep("Normal(0,5)",nn))) %>% 
  arrange(slope_prior,slope) %>% 
  mutate(draw = rep(1:nn,times = 3))

# calculating the expected counts for each 
priors_changes <- priors_changes %>%
  expand_grid(.,forest_range) %>% #this replicates the prior_changes dataframe for every level of forest_range
  mutate(expect_prop = exp(intercept + slope*proportion_forest),
         expect_centered = exp(intercept + slope*pforest_centered),
         expect_scaled = exp(intercept + slope*pforest_scaled)) %>% 
  arrange(slope_prior,draw)


prior_slopes1 <- ggplot(data = priors_changes)+
  geom_line(aes(x = proportion_forest,y = expect_prop,
                group = draw,
                colour = slope_prior))+
  facet_grid(rows = vars(slope_prior),
             scales = "free_y")+
  scale_y_continuous(trans = "log10", labels = comma)+
  labs(title = "pforest")+
  ylab("Expected count")

prior_slopes2 <- ggplot(data = priors_changes)+
  geom_line(aes(x = proportion_forest,y = expect_centered,
                group = draw,
                colour = slope_prior))+
  facet_grid(rows = vars(slope_prior),
             scales = "free_y")+
  scale_y_continuous(trans = "log10", labels = comma)+
  labs(title = "centered")+
  ylab("Expected count")

prior_slopes3 <- ggplot(data = priors_changes)+
  geom_line(aes(x = proportion_forest,y = expect_scaled,
                group = draw,
                colour = slope_prior))+
  facet_grid(rows = vars(slope_prior),
             scales = "free_y")+
  scale_y_continuous(trans = "log10", labels = comma)+
  labs(title = "scaled")+
  ylab("Expected count")

print(prior_slopes3 + prior_slopes1 + prior_slopes2 + 
        plot_layout(guides = "collect"))


```


So you can see that the standard normal prior `normal(0,1)` does well with the scaled predictor values `scale(proportion_forest)`, but the other priors generate ridiculous estimates of abundance using the same predictor. By contrast, for the raw proportion forest or the centered proportion forest, the 




## More formal prior predictions

As we've done a few times in the course material you can also explore the priors by first building a model in Stan or `ulam()` and then sampling from the full prior predictive distribution integrating information across all parameters in the model.

```{r, ulam_model, message=FALSE,warning=FALSE, cache=TRUE}


#partial pooling varying slopes and intercept model with NB distribution
m1<- ulam(
  alist(
    count ~ dgampois( lambda , phi ),
    log(lambda) <- sp[species_ind] + b_for[species_ind]*pforest_scale,
    sp[species_ind] ~ dnorm( 0 , 5 ), #using the wide, zero-mean normal prior from above
    b_for[species_ind] ~ dnorm( b_for_bar , sigma_b_for ),
    b_for_bar ~ dnorm( 0 , 5 ),
    sigma_b_for ~ dexp(1),
    phi ~ dexp(1)
  ), data=birds_sub , chains=4, cores = 4,
  cmdstan = TRUE)

```


```{r ulam_prior}
#prior simulation
prior_m1 <- extract.prior(m1)


```







